---
title: "IR bacteria vs P. infestans  "
author: "Philemon"
date: "2026-02-18"
output: html_document
---

```{r}

library(readxl)
library(dplyr)
library(glmmTMB)
library(emmeans)
library(DHARMa)
library(ggplot2)
#install.packages("writexl")
library (writexl)
```


```{r}
# 1) Load + standardize column names
df_raw <- read_excel("~/Ad_Planta_Trials/Raw_data/IR_SelectedbacteriavsPinfestans_Rep One and Two.xlsx") %>%
  rename(
    Treatment = Treatment,
    Rep = Replicate,
    severity_leaf = Diseased_leaf_area
  ) %>%
  mutate(
    Treatment = as.factor(Treatment),
    Rep = as.factor(Rep),
    Plant = as.factor(Plant),
    Leaf  = as.factor(Leaf),
    severity_leaf = as.numeric(severity_leaf)
  )
View(df_raw)

```


```{r}

# 2) Plant is the experimental unit: average leaves within plant
df_plant <- df_raw %>%
  group_by(Rep, Treatment, Plant) %>%
  summarise(severity = mean(severity_leaf, na.rm = TRUE), .groups = "drop")

# 3) Prepare for Gamma (no zeros). Offset should be small.
df_plant <- df_plant %>%
  mutate(severity_pos = severity + 0.5)

# 4) Set reference level for comparisons (infected control)
df_plant <- df_plant %>%
  mutate(Treatment = relevel(Treatment, ref = "Non-treated control"))

# 5) Fit mixed model: Treatment fixed, Rep random
m1 <- glmmTMB(
  severity_pos ~ Treatment + (1|Rep),
  family = Gamma(link = "log"),
  data = df_plant
)
summary(m1)


# 6) Diagnostics
res <- simulateResiduals(m1, n = 1000)
plot(res)

#Dagnostics show that Residuals change with the fitted valuesariance changes with the mean in a way the model doesn’t handle,there’s heterogeneity by Treatment/Rep,missing random structure
#Fix: allow different dispersion per Treatment
m2 <- glmmTMB(
  severity_pos ~ Treatment + (1|Rep),
  family = Gamma(link="log"),
  dispformula = ~ Treatment,
  data = df_plant
)

#Beta Distribution: If your data is strictly between 0 and 1 (or 0% and 100%), the Beta distribution is usually much better than Gamma or Tweedie.
df_plant$severity_pos_scaled <- df_plant$severity_pos / 100

m2 <- glmmTMB(
  severity_pos_scaled ~ Treatment + (1|Rep),
  family = beta_family(link="logit"),
  dispformula = ~ Treatment,
  data = df_plant
)
summary(m2)
res2 <- simulateResiduals(m2, n=1000)
plot(res2)


# 7) Estimated marginal means
emm <- emmeans(m2, ~ Treatment)

# 8) Dunnett-style: each treatment vs non-treated control 
contr_pos <- contrast(
  emm,
  method = "trt.vs.ctrl",
  ref = "Non-treated control",
  adjust = "BH"
)
# On model scale
summary(contr_pos)

#Scientific question is: “Which treatments differ from the diseased control?”
#Then the best-matched procedure is Dunnett (many-to-one). It’s more powerful than BH because it’s designed specifically for #comparisons vs a single control.

contr_pos <- contrast(
  emm,
  method = "trt.vs.ctrl",
  ref = "Non-treated control",
  adjust = "dunnettx"   # if supported
)
summary(contr_pos)
summary(contr_pos, type="response")

# On response scale (ratios = treated / control)
contr_pos_resp <- summary(contr_pos, type = "response", infer = c(TRUE, TRUE)) %>%
  as.data.frame()

contr_pos_resp

```
Convert ratios to “% reduction in severity” (and CI)
```{r}

library(tidyverse)

# 1. Identify the ratio and CI columns dynamically
# Identify the raw columns from the summary object
# Based on emmeans screenshot, these are the exact names
# 1. Force the use of the original decimal values from the summary object
# We use contr_pos_resp because it is the original data frame before mutate errors
raw_ratios <- contr_pos_resp$odds.ratio
raw_low    <- contr_pos_resp$asymp.LCL
raw_high   <- contr_pos_resp$asymp.UCL

# 2. Re-calculate eff_plot from scratch
eff_plot_final <- contr_pos_resp %>%
  mutate(
    Treatment_raw = gsub(" / \\(Non-treated control\\)", "", contrast),
    Treatment_raw = trimws(Treatment_raw),
    
    # Force the math: (1 - 0.069) * 100 = 93.1
    perc_reduction = (1 - raw_ratios) * 100,
    
    # Flip CIs: 1 - High Ratio = Lower Reduction
    perc_low  = (1 - raw_high) * 100,
    perc_high = (1 - raw_low)  * 100,
    
    # Re-apply labels and factors
    Treatment_lab = dplyr::recode(Treatment_raw, !!!treat_labels, .default = Treatment_raw),
    Treatment_lab = factor(Treatment_lab, levels = custom_sequence),
    
    p_label = ifelse(p.value < 0.001, "p<0.001", paste0("p=", signif(p.value, 2)))
  ) %>%
  filter(!is.na(Treatment_lab))

# 3. Check the data! 
# Run this line: If perc_reduction is still > 100, your odds.ratio is wrong.
print(head(eff_plot_final[c("Treatment_lab", "perc_reduction")]))
View (eff_plot_final)

#Save the percentage disease reduction summary
write_xlsx(eff_plot_final, "~/Ad_Planta_Trials/Outputfiles/Disease_Reduction_Summary_IR.xlsx")


p_efficacy <- ggplot(eff_plot_final, aes(x = Treatment_lab, y = perc_reduction)) +
  # Baseline reference at 0
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  
  # Bar layer
  geom_col(width = 0.65, fill = "steelblue", color = "black") + 
  
  # Error bar layer - Modified to show only the top side
  geom_errorbar(aes(ymin = perc_reduction, ymax = perc_high), 
                width = 0.2, 
                linewidth = 0.6) + 
  
  # Significance labels
  geom_text(aes(y = pmin(perc_high + 4, 105), label = p_label), 
            size = 3, vjust = 0) +
  
  # Formatting
  scale_y_continuous(limits = c(0, 115), breaks = seq(0, 100, 20), expand = c(0, 0)) +
  labs(
    x = "Treatment", 
    y = "Disease Reduction vs. Non-treated control (%)"
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 10, angle = 45, vjust = 1, hjust = 1, face = "italic"),
    panel.grid.minor = element_blank()
  )

print(p_efficacy)

ggsave(
  filename = "Disease_Reduction_Summary_IR.tiff",
  plot = p_efficacy,
  path = "~/Ad_Planta_Trials/Outputfiles/Plots",
  width = 7.5, height = 6.5, dpi = 300, units = "in",
  device = "tiff", compression = "lzw"
)








write.csv(eff_df, "~/Ad_Planta_Trials/Outputfiles/eff_df_table_IR.csv", row.names = FALSE)


```

Plot: % reduction vs infected control

```{r}


treat_labels <- c(
  "Nutrient broth" = "Nutrient broth",
  "Bacillus sp._109" = "Bacillus sp.Pb109", # Fixed mismatch: removed 'Pb' from key
  "Bacillus sp._Pb168" = "Bacillus sp.Pb168",
  "Microbacterium sp._Pb539" = "Microbacterium sp.Pb539",
  "Paenibacillus sp._Pb355 " = "Paenibacillus sp.Pb355 ",
  "Pseudarthrobacter sp._Pb177" = "Pseudarthrobacter sp.Pb177",
  "Pseudomonas sp._Pb114" = "Pseudomonas sp.Pb114",
  "Pseudomonas sp._Pb16" = "Pseudomonas sp.Pb16",
  "Pseudomonas sp._Pb170" = "Pseudomonas sp.Pb170",
  "Pseudomonas sp._Pb242" = "Pseudomonas sp.Pb242",
  "Pseudomonas sp._Pb271" = "Pseudomonas sp.Pb271",
  "Pseudomonas sp._Pb285" = "Pseudomonas sp.Pb285",
  "Pseudomonas sp._Pb457" = "Pseudomonas sp.Pb457",
  "Pseudomonas sp._Pb460" = "Pseudomonas sp.Pb460",
  "Sphingomonas sp._Pb171" = "Sphingomonas sp.Pb171",
  "Streptomyces sp._Pb39" = "Streptomyces sp.Pb39"
)

# The specific order you want on the X-axis
custom_sequence <- c(
  "Nutrient broth",
  "Bacillus sp.Pb109", "Bacillus sp.Pb168",
  "Microbacterium sp.Pb539",
  "Paenibacillus sp.Pb355 ",
  "Pseudarthrobacter sp.Pb177",
  "Pseudomonas sp.Pb16", "Pseudomonas sp.Pb114", "Pseudomonas sp.Pb170", 
  "Pseudomonas sp.Pb242", "Pseudomonas sp.Pb271", "Pseudomonas sp.Pb285", 
  "Pseudomonas sp.Pb457", "Pseudomonas sp.Pb460",
  "Sphingomonas sp.Pb171",
  "Streptomyces sp.Pb39"
)

# Detect columns dynamically to avoid 'Object not found' errors
resp_col <- intersect(names(contr_pos_resp), c("ratio", "odds.ratio", "response"))[1]
low_col  <- intersect(names(contr_pos_resp), c("asymp.LCL", "lower.CL", "conf.low"))[1]
high_col <- intersect(names(contr_pos_resp), c("asymp.UCL", "upper.CL", "conf.high"))[1]

eff_plot <- contr_pos_resp %>%
  mutate(
    # Clean the Treatment string from the contrast output
    Treatment_raw = gsub(" / Non-treated control| - Non-treated control", "", contrast),
    Treatment_raw = trimws(Treatment_raw),
    
    # Map to pretty labels
    Treatment_lab = dplyr::recode(Treatment_raw, !!!treat_labels, .default = Treatment_raw),
    
    # Set display order using your custom sequence
    Treatment_lab = factor(Treatment_lab, levels = custom_sequence),
    
    # Transform CI: High Ratio = Low Reduction
    perc_reduction = (1 - .data[[resp_col]]) * 100,
    perc_low = (1 - .data[[high_col]]) * 100,
    perc_high = (1 - .data[[low_col]]) * 100,
    
    p_label = ifelse(p.value < 0.001, "p<0.001", paste0("p=", signif(p.value, 2)))
  ) %>%
  filter(!is.na(Treatment_lab))


p_bar_v <- ggplot(eff_plot, aes(x = Treatment_lab, y = perc_reduction)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_col(width = 0.6, fill = "steelblue", color = "black") + 
  geom_errorbar(aes(ymin = perc_low, ymax = perc_high), width = 0.2) + 
  geom_text(aes(y = pmin(perc_high + 5, 103), label = p_label), 
            size = 3.2, vjust = 0) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 20), expand = c(0, 0)) +
  labs(x = "Treatment", y = "Disease reduction vs Non-treated control (%)") +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 11, angle = 45, vjust = 1, hjust = 1, face = "italic"),
    panel.grid.minor = element_blank()
  )

p_bar_v










# 3) Prepare plot data + p-value labels
library(dplyr)
library(ggplot2)

eff_plot <- eff_df %>%
  filter(Treatment != "Non-treated control") %>%
  mutate(
    # Map the raw names (with underscores) to your clean labels
    # .default = Treatment ensures names are kept if they aren't in the list
    Treatment_lab = dplyr::recode(Treatment, !!!treat_labels, .default = Treatment),
    
    # Set the order using your custom_sequence
    Treatment_lab = factor(Treatment_lab, levels = custom_sequence),
    
    # Standardize p-value labels
    p_label = ifelse(p.value < 0.001, "p<0.001", paste0("p=", signif(p.value, 2)))
  ) %>%
  # Crucial: Remove rows where Treatment_lab is NA to prevent empty plots
  filter(!is.na(Treatment_lab))


p_bar_v <- ggplot(eff_plot, aes(x = Treatment_lab, y = perc_reduction)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_col(width = 0.6, fill = "steelblue", color = "black") + 
  # Use the CI columns you created in your previous mutate step
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2, linewidth = 0.6) + 
  geom_text(aes(y = pmin(asymp.UCL + 5, 103), label = p_label), 
            size = 3.2, vjust = 0) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 20), expand = c(0, 0)) +
  labs(x = "Treatment", y = "Disease reduction vs Non-treated control (%)") +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 10, angle = 45, vjust = 1, hjust = 1, face = "italic"),
    panel.grid.minor = element_blank()
  )

print(p_bar_v)













