---
title: "Tomicro2025_16S_statistics_ordination"
author: "Philemon Orwa"
date: "2025-02-14"
output: html_document
---

### Analysis alpha-diversity with rarefied dataset
Load data
```{r}
my_asv_table <- read.table("~/Tomicro_2025/16S_Tomicro_2025_Rarefied_Table_1000_times_Clean_ASVnumbers_2.txt", header=T, sep="\t", dec=".", fill=FALSE) #add manually in textfile a 0 in front of single numbers of samples 1, 2, 3, 4 ,5 ...
head(my_asv_table)
dim(my_asv_table)

metadat <- read.table("~/Tomicro_2025/Metadata_02_2025.txt", header = TRUE, row.names = 2, sep = "\t", quote = "\"")
View(metadat)
colnames(metadat)  # Check for column names
colnames(metadat) <- trimws(colnames(metadat))  # Remove spaces from column names
metadat$Plant.condition <- factor(trimws(metadat$Plant.condition))  # Remove extra spaces
metadat$Soil.origin <- factor(trimws(metadat$Soil.origin))  # Remove extra spaces
metadat$Microcompartment <- factor(trimws(metadat$Microcompartment))  # Remove extra spaces
metadat$Replicate <- factor(trimws(metadat$Replicate))  # Remove extra spaces

stopifnot(identical(rownames(my_asv_table), rownames(metadat)))


```

Richness
```{r}
# Define function for richness
Species.richness <- function(x) {
  sum(x > 0.5)
}

alpha_divs <- metadat
alpha_divs$richness <- apply(my_asv_table, 1, Species.richness)

View(alpha_divs)

#Fit ANOVA model (alpha-diversity)
anova_model <- aov(
  richness ~ Soil.origin * Plant.condition * Microcompartment,
  data = alpha_divs
)

#Check ANOVA assumptionsp > 0.05 → residuals are approximately normal → ANOVA valid p ≤ 0.05 → normality violated → do NOT use ANOVA

shapiro.test(residuals(anova_model))

#Shapiro-Wilk normality test data:  residuals(anova_model) W = 0.95819, p-value = 0.08527

qqnorm(residuals(anova_model))
qqline(residuals(anova_model), col = "red")

#Run ANOVA (assumptions are met)

summary(anova_model)


ggplot(alpha_divs, aes(y = richness, x = Soil.origin, col = Plant.condition)) +
  geom_boxplot() +
  facet_grid(~Microcompartment) +
  stat_compare_means(method = "anova", label = "p.format")

TukeyHSD(anova_model)
TukeyHSD(anova_model, which = "Plant.condition")


```

Shannon index
Shannon or Shannon–Weaver (or Shannon–Wiener) index is defined as H = -sum p_i log(b) p_i, where p_i is the proportional abundance of species i and b is the base of the logarithm. It is most popular to use natural logarithms, but some argue for base b = 2 (which makes sense, but no real difference).
```{r}
alpha_divs$shannon <-diversity(my_asv_table, index = "shannon", base = exp(1), MARGIN = 1)
View(alpha_divs)

#Fit ANOVA model (alpha-diversity)
anova_model <- aov(
  shannon ~ Soil.origin * Plant.condition * Microcompartment,
  data = alpha_divs
)


qqnorm(residuals(anova_model))
qqline(residuals(anova_model), col = "red")

ggplot(data=alpha_divs, aes(y=shannon, x=Soil.origin, col = Plant.condition))+ geom_boxplot() + facet_grid(~Microcompartment) + stat_compare_means(method = "anova", label = "p.format")


```

Evenness
```{r}
diversity <- diversity(my_asv_table) #default=shannon
speciesrichness <-specnumber(my_asv_table)
alpha_divs$evenness<- diversity/log(speciesrichness)

#Fit ANOVA model (alpha-diversity)
anova_model <- aov(
  evenness ~ Soil.origin * Plant.condition * Microcompartment,
  data = alpha_divs
)

qqnorm(residuals(anova_model))
qqline(residuals(anova_model), col = "red")


ggplot(data=alpha_divs, aes(y=evenness, x=Soil.origin, col = Plant.condition))+ geom_boxplot() + facet_grid(~Microcompartment)+ stat_compare_means(method = "anova", label = "p.format")
```

Simpson index
Both variants of Simpson's index are based on D = sum p_i^2. Choice simpson returns 1-D and invsimpson returns 1/D.
```{r}
alpha_divs$simpson <- diversity(my_asv_table, index = "simpson", base = exp(1), MARGIN = 1)

anova_model <- aov(
  simpson ~ Soil.origin * Plant.condition * Microcompartment,
  data = alpha_divs
)

shapiro.test(residuals(anova_model))

qqnorm(residuals(anova_model))
qqline(residuals(anova_model), col = "red")



ggplot(data=alpha_divs, aes(y=simpson, x=Soil.origin, col = Plant.condition))+ geom_boxplot() + facet_grid(~Microcompartment)+ stat_compare_means(method = "anova", label = "p.format")
```

#ANOVA/statistics on mean of diversity indices
Behr et al. 2023: Significant differences in a-diversity (ASV Richness, Simpson, Shannon Index, and Evenness) were estimated with two-way ANOVA tests. If the data failed to fulfill the normality criteria based on the Shapiro test, the non-parametric aligned-rank ANOVA was performed (“ARTool” package, v.0.11.1, Kay et al., 2021).

3-way-ANOVA Richness 
```{r}
library(car)
library(rcompanion)

model_1 <- aov(richness ~ Soil.origin*Plant.condition*Microcompartment, alpha_divs)
anova(model_1) 
qqnorm(residuals(model_1))
leveneTest(model_1) 

#run Tukey’s test for pairwise comparisons
TukeyHSD(model_1)

```

```{r}
model_2 <- aov(shannon ~ Soil.origin*Plant.condition*Microcompartment, alpha_divs)
anova(model_2) 

qqnorm(residuals(model_2))
leveneTest(model_2) 

#run Tukey’s test for pairwise comparisons
TukeyHSD(model_2)


```

```{r}
model_3 <- aov(simpson ~ Soil.origin*Plant.condition*Microcompartment, alpha_divs)
anova(model_3) 

qqnorm(residuals(model_3))
leveneTest(model_3) 


#run Tukey’s test for pairwise comparisons
TukeyHSD(model_3)
```

```{r}
model_4 <- aov(evenness ~ Soil.origin*Plant.condition*Microcompartment, alpha_divs)
anova(model_4) 

qqnorm(residuals(model_4))
leveneTest(model_4) 

#run Tukey’s test for pairwise comparisons
TukeyHSD(model_4)
```












