---
title: "Tomicro2025_16S_Pre-processing"
author: "Philemon Orwa"
date: "2025-02-11"
output: html_document
---

```{r}
library(ggplot2) #ggplot2: Fancy graphs
library(ggpubr) #ggplot expansion for more fancy graphs
library(readxl)
library(readr)
library(dplyr)# Data handling
library(vegan)# Multivariate stat tool
library(tidyr) # Data handling
library(RColorBrewer) # Colours for fancy graphs
library(tibble)# Data handling: rownames/columnanmes
#install.packages("phyloseq")
#library(phyloseq) # tool for 16S amplicon analysis
library(lme4) # mixed effect models 
library(forcats)
library(tidyverse)
library(tibble)

```

Read data
```{r}
ASVs <- read.csv("~/Tomicro_2025/ASVtable2_16S_Tomicro_2025.csv", header = TRUE)
#add manually in textfile a 0 in front of single numbers 1, 2, 3, 4 ,5 ...
head(ASVs)
dim(ASVs)

tax <- read.csv("~/Tomicro_2025/taxa_16S_Tomicro_2025.csv", header = TRUE) 
head(tax)
dim(tax)

```

```{r}

# Assign human-friendly ASV IDs
ASVs <- ASVs %>%
  mutate(ASVid = paste0("ASV", row_number()))

tax <- tax %>%
  mutate(ASVid = paste0("ASV", row_number()))

# Merge ASV abundance table with taxonomy using ASVid
tax_ASV <- inner_join(ASVs, tax, by = "ASVid")

# View structure
glimpse(tax_ASV)

write.csv(tax_ASV, "tax_ASV_16S_Tomicro_2025.csv")

tax_ASV <- tax_ASV %>%
    select(-X.x, -X.y)  # Remove unnecessary columns

tax_ASV <- tax_ASV %>%
  relocate(ASVid, .before = everything())  # Moves ASVid to the first column
head(tax_ASV)
dim(tax_ASV)

tax_ASV_filt <- tax_ASV %>%
  filter(Kingdom == "Bacteria") %>%  # Step 1: Keep only Bacteria
  filter(!str_detect(Order, "Mitochondria|Chloroplast")) %>%  # Step 2: Remove rows where Order contains "Mitochondria" or "Chloroplast"
  filter(!str_detect(Family, "Mitochondria|Chloroplast"))  # Step 3: Remove rows where Family contains "Mitochondria" or "Chloroplast"

head(tax_ASV_filt)
dim(tax_ASV_filt)

write.csv(tax_ASV_filt, "tax_ASV_fil_16S_Tomicro_2025.csv")

# Check if filtering worked
unique(tax_ASV$Kingdom)  # Before filtering
unique(tax_ASV_filt$Kingdom)  # After filtering

colnames(tax_ASV_filt)

sum(duplicated(tax_ASV_filt$ASVid)) 

ASV_subset <- tax_ASV_filt[,c(1:49)]
dim(ASV_subset)
head(ASV_subset)

rownames(ASV_subset) <- ASV_subset[, 1]
ASV_subset <- ASV_subset[, -1]

head(ASV_subset)




# Identify the last sample column (before Kingdom appears)
num_sample_cols <- which(colnames(tax_ASV_filt) == "Kingdom") - 1

# Subset only the ASV abundance data
ASV_subset <- tax_ASV_filt[, c(2:num_sample_cols)]  # Excludes ASVid and taxonomy
dim(ASV_subset)
head(ASV_subset)
write.csv(ASV_subset, "ASV_subset_fil_16S_Tomicro_2025.csv")

sum(duplicated(ASV_subset[, 1]))  # Count duplicates
unique(ASV_subset[, 1])  # View unique values

ASV_subset

rownames(ASV_subset) <- ASV_subset[, 1]  # Set ASVid as rownames
ASV_subset <- ASV_subset[, -1]  # Remove ASVid column

# Ensure samples are sorted alphabetically
ASV_subset <- ASV_subset[, order(names(ASV_subset))]

# View final dataset
View(ASV_subset)

ASV_subset <- ASV_subset[ , order(names(ASV_subset))]
View(ASV_subset)

ASVs_clean <- ASV_subset[-which(rowSums(ASV_subset[sapply(ASV_subset, is.numeric)]) <= 5), ]
dim(ASVs_clean)
head(ASVs_clean)

#Exclude ASVs with all 0 in one row
ASVs_no0 <- ASVs_clean[rowSums(ASVs_clean[, -1] > 0) != 0, ]
dim(ASVs_no0)
head(ASVs_no0)
View(ASVs_no0)


#Export
write.table(ASVs_no0, "~/Tomicro_2025/ASV_table_16S_Tomicro_2025_no0_ASVNumbers_2.txt", dec = ".", sep = "\t")

#Clean taxonomy dataset
# Convert row names to a column 
ASVs_no0_2 <- ASVs_no0 %>%
  rownames_to_column(var = "rowname")
head(ASVs_no0_2)
dim(ASVs_no0_2)
View(ASVs_no0_2)

#change name of first column
colnames(tax_ASV_filt)[1] <- "rowname"
head(tax_ASV_filt)
View(tax_ASV_filt)
#select columns in taxonomy
rowname <- tax_ASV_filt[, 1]
last_col <- tax_ASV_filt[, (ncol(tax_ASV_filt)-5):ncol(tax_ASV_filt)]
tax_new <- cbind(rowname, last_col)
head(tax_new)
dim(tax_new)


#join datasets
tax_clean <- inner_join(ASVs_no0_2, tax_new, by="rowname")
dim(tax_clean)
View(tax_clean)

#select taxonomy columns
ASVid <- tax_clean[, 1]
last_cols <- tax_clean[, (ncol(tax_clean)-5):ncol(tax_clean)]
tax_final <- cbind(ASVid, last_cols)
dim(tax_final)
head(tax_final)
View(tax_final)
#Export
write.table(tax_final, "~/Tomicro_2025/16S_Tomicro_2025_Table_SILVAtaxonomy_CLEAN_2.txt", dec = ".", sep = "\t")


```

#Create rarefied dataset
Read data
```{r}
ASVs_no0_new <- read.table("~/Tomicro_2025/ASV_table_16S_Tomicro_2025_no0_ASVNumbers_2.txt", header=T, sep="\t", dec=".", fill=FALSE) 
head(ASVs_no0_new)
dim(ASVs_no0_new)
```

transpose data
```{r}
#transpose data
tASVs<- t(ASVs_no0_new) # samples in rows
dim(tASVs)
head(tASVs)
str(tASVs)
View(tASVs)
```

Visualization of rarefaction curve 1000 random rarefactions
```{r}
library(vegan)
myrare <- rarecurve(tASVs, step = 1000, cex=0.4)
```

get min rowSum 
```{r}
rowASVs <- rowSums(tASVs)
min(rowASVs)
```

creating 1000x rarefied table for further analysis of alpha-diversity
```{r}
Have_you_Runned_previously_RF<-"NO"
if (Have_you_Runned_previously_RF=="NO"){
  list_of_1000_rarefaction=lapply(1:1000, function(i){ rrarefy(tASVs, 14866)})
}

mean.dat <- as.data.frame(apply(simplify2array(lapply(list_of_1000_rarefaction, as.matrix)),1:2,mean))
View(mean.dat)
dim(mean.dat)

write.table(mean.dat, file="~/Tomicro_2025/16S_Tomicro_2025_Rarefied_Table_1000_times_Clean_ASVnumbers_2.txt", sep = "\t")

```
To account for uneven sequencing depth, we repeatedly performed 1000 rarefactions to our dataset’s lowest number of sequences. Average ASV abundances were calculated based on the 1000 rarefactions.



