---
title: "Tomicro2025_16S_square_barplots"
author: "Philemon"
date: "2025-07-11"
output: html_document
---
```{r}
library(tidyverse)
library(readxl)
library(stringr)
library(dplyr)
library(ggtext) # we can change fonts in plots
library(RColorBrewer)
library(ggplot2)
```
Load clean datasets

```{r}

metadata <- read.table("~/Tomicro_2025/Metadata_02_2025.txt", header=T, sep="\t", dec=".") %>%
  select(ID, Plant.condition, Microcompartment, Soil.origin, Compsoil, CompSoilDis) 
View(metadata)

#rarefied dataset (according to Schloss 2024)
ASV_counts <- read.table("~/Tomicro_2025/16S_Tomicro_2025_Rarefied_Table_1000_times_Clean_ASVnumbers_2.txt", header=T, sep="\t", dec=".", fill=FALSE)#this file was cleaned for mitos, chloros and all taxa which are not bacteria, also it is a rarefied dataset
dim(ASV_counts)
View(ASV_counts)
#change column name
    ASV_counts <- as.data.frame(ASV_counts)
    ASV_counts <- rownames_to_column(ASV_counts, var = "ID")
    View(ASV_counts)
#continue to long format
    ASV_counts <- ASV_counts %>%
    pivot_longer(-ID, names_to="ASV", values_to = "count")
    View(ASV_counts)

tax_ASV <- read.table("~/Tomicro_2025/16S_Tomicro_2025_Table_SILVAtaxonomy_CLEAN_2.txt", header=TRUE, sep="\t") 
#rename column
    colnames(tax_ASV)[which(names(tax_ASV) == "ASVid")] <- "ASV"

View(tax_ASV)
dim(tax_ASV)

```
Combine in one dataset
```{r}
ASV_relabu <- dplyr::inner_join(metadata, ASV_counts, by = "ID") %>%
  dplyr::inner_join(., tax_ASV, by = "ASV") %>%
  group_by(ID) %>% mutate(rel_abund = count / sum(count)) %>%
  ungroup() %>%
  select(-count) 

ASV_relabu
dim(ASV_relabu)

#check if all values add up to 1
ASV_relabu %>%
  group_by(ID) %>%
  summarize(total = sum(rel_abund))
#check if everything apart from bacteria is out
unique(ASV_relabu$Kingdom)
```
```{r}

```{r}

#Long format
ASV_relabu_long <- ASV_relabu %>%
  pivot_longer(c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "ASV"),
               names_to="level",
               values_to="taxon")

ASV_relabu_long

```


Square barplot with standard deviation

```{r}
taxon_ASV_relabu <- ASV_relabu_long %>%
  filter(level=="Phylum") %>% #change here taxon level
  group_by(Soil.origin, Microcompartment, ID, taxon) %>%
  summarize(rel_abund = sum(rel_abund), .groups="drop") %>%
  mutate(taxon = ifelse(is.na(taxon), "Unclassified", taxon)) %>%
  group_by(Soil.origin, Microcompartment, taxon) %>%
  summarize(mean_rel_abund = 100*mean(rel_abund),  sd = 100 * sd(rel_abund), .groups = "drop") %>%   #100* makes relabu in %, We calculate the standard deviation of the mean sd(rel_abund) after calculating the mean relative abundance
  mutate(taxon = str_replace_all(taxon, "\\*", ""))
  #mutate(taxon = ifelse(taxon %in% c("Other", "Unclassified"),
                      #taxon,
                      #paste0("*", taxon, "*")))
#check highest abundances  (check percentage you want as limit)
taxon_ASV_relabu %>%
  group_by(taxon)%>%
  summarize(max = max(mean_rel_abund))%>%
  arrange(desc(max))%>%
  slice_head(n = 10)  # Top 10
  
#pool low abundant phyla 
taxon_pool <- taxon_ASV_relabu %>%
  group_by(taxon)%>%
  summarize(pool = max(mean_rel_abund)<5, #add here percentage you want as limit
            mean=mean(mean_rel_abund),
            .groups= "drop")

taxon_combine <- dplyr::inner_join(taxon_ASV_relabu, taxon_pool, by="taxon")%>%
  mutate(taxon=if_else(pool, "Other", taxon))%>%
  group_by(Soil.origin, Microcompartment, taxon)%>%
  summarize(mean_rel_abund = sum(mean_rel_abund),
            sd = sum(sd),
            mean = min(mean),
            .groups="drop")%>%
  mutate(taxon=factor(taxon),
         taxon=fct_reorder(taxon, mean, .desc=FALSE))
taxon_combine

#debug
unique(taxon_combine$Plant.condition)




Fig_Phylum <- taxon_combine %>%
  ggplot(aes(fill=Soil.origin, x=mean_rel_abund, y=taxon)) +
  geom_errorbar(aes(xmin=mean_rel_abund, xmax=mean_rel_abund + sd),
                position = position_dodge2(width = 0.8, reverse = TRUE, padding = 0),
                width = 0.8,
                size = 0.2) +
  geom_col(position = position_dodge2(width = 0.8, reverse = TRUE, padding = 0), width=0.8) +
  scale_fill_manual(name=NULL, breaks=c("SoilA", "SoilB"),
                   labels=c("Soil A", "SOil B"), 
                   values=c("#ffbf00", "#5b9bd5")) + # yellow #ffbf00, blue #5b9bd5
   scale_x_continuous(expand=c(0, 0)) +
  labs(y=NULL,
       x="Mean Relative Abundance (%)") +
  facet_grid(~Microcompartment, scale = "free_y", switch = "y")+
  theme_classic() +
  theme(axis.text.y = ggtext::element_markdown(size = 12, color = "black"),  #  y-axis label font
        legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        legend.position = c(0.95, 0.6),
        panel.grid.major.x = element_line(color="black", size = 0.25),
        panel.spacing = unit(1, "lines"))

Fig_Phylum


# Save plot as TIFF
ggsave(
  filename = "Fig_Phylum_soil_origin.jpeg",
  plot = Fig_Phylum,
  path = "~/Tomicro_2025/Square Bar plots/Soil origin",
  width = 8, height = 5, dpi = 300, units = "in",
  device = "jpeg"
)

#ggsave("Philemon.tiff", width=8, height=5)
```

